---
title: "Gene Ontology Enrichment Analysis for C. albicans Infection Time Course"
author: "M. LeGro"
date: "2026-07-22"
output:
  html_document:
    highlight: tango
    number_sections: yes
    theme: united
    toc: yes
    toc_depth: 4
  pdf_document:
    number_sections: yes
    toc: yes
  word_document:
    toc: yes
editor_options:
  chunk_output_type: inline
---

# Introduction

After the differential expression analysis with limma-voom (each infected time point -- 6hpi, 1dpi, 3dpi, 5dpi, 7dpi -- contrasted against uninfected Controls), GO terms are mapped to the gene rankings to look at pathway-level expression as a whole. This uses the same clusterProfiler-based pipeline developed for the Rad51(RNAi) project (an update from the classic topGO pipeline used previously), for both over-representation analysis (ORA) and rank-based gene set enrichment analysis (GSEA). Since this is the same *S. mediterranea* genome/annotation used across all of these dissertation chapters, the same gene ID-to-GO mapping files are reused here rather than duplicated per project.

## Analysis set up and converting gene ID and GO files to analysis-compatible files

```{r, warning=FALSE, comment=NA, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
library(data.table)
library(dplyr)
library(stringr)
library(clusterProfiler)
library(GO.db)
library(enrichplot)
library(AnnotationDbi)
library(ggplot2)

setwd("/Volumes/PortableSSD/CA_infectionproj/")

# Shared reference location (same genome/annotation/GO mapping files used across the
# PTEN, Rad51, and this project) rather than a copy kept in each project's own folder.
reference_dir <- "/Volumes/PortableSSD/Rad51/"

de_analysis_date <- "2026.07.21"   # date used when the DE analysis Rmd wrote its toptable CSVs -- update if yours differs
go_analysis_date <- "2026.07.22"   # date label for this GO analysis's own output files

timepoints <- c("6hpi", "1dpi", "3dpi", "5dpi", "7dpi")

# Import the GO-to-gene ID files (unchanged from the Rad51 pipeline -- these describe
# the genome/annotation, not any particular RNAi or infection experiment)
file_map_simple <- read.delim(file.path(reference_dir, "2018.5.7_geneID2GO.txt"))  # 2-col file: Contig [tab] Goterm
file_map_rich   <- read.delim(file.path(reference_dir, "GOterms_dd_Smed_v6.txt"))

# Helper: normalize names to lowercase, underscores, no punctuation
clean_names <- function(x) tolower(gsub("[^a-z0-9]+", "_", trimws(x)))

# --- SIMPLE MAP ---
map_simple <- file_map_simple
names(map_simple) <- clean_names(names(map_simple))

# --- RICH MAP ---
map_rich <- file_map_rich
names(map_rich) <- clean_names(names(map_rich))
names(map_rich) <- c("Contig", "Goterm", "GO_name", "GO_desc")

# --- Clean values ---
clean_go_id <- function(x) {
  x <- toupper(trimws(x))
  x
}

map_rich <- map_rich |>
  mutate(Contig = trimws(Contig),
         Goterm = clean_go_id(Goterm),
         GO_name = trimws(GO_name),
         GO_desc = trimws(GO_desc)) |>
  filter(Contig != "", Goterm != "") |>
  distinct()

# --- Build TERM2GENE & TERM2NAME ---
term2gene <- data.frame(term = file_map_simple$Goterm, gene = file_map_simple$Contig, stringsAsFactors = FALSE)

# TERM2NAME (pretty labels)
u_terms <- unique(term2gene$term)
ann <- AnnotationDbi::select(GO.db, keys = u_terms, columns = "TERM", keytype = "GOID")
term2name <- data.frame(term = ann$GOID, name = ann$TERM, stringsAsFactors = FALSE)

na_idx <- is.na(term2name$name)
term2name$name[na_idx] <- term2name$term[na_idx]

# confirm structure
stopifnot(all(c("term", "gene") %in% names(term2gene)))
stopifnot(all(grepl("^GO:\\d+$", term2gene$term)))
cat("Unique contigs:", length(unique(term2gene$gene)),
    " Unique GO terms:", length(unique(term2gene$term)),
    " Duplicated rows (should be 0):", sum(duplicated(term2gene)), "\n")

# create BP-only mappings
u_terms <- unique(term2gene$term)
ann     <- AnnotationDbi::select(GO.db, keys = u_terms, columns = "ONTOLOGY", keytype = "GOID")
term2ont <- data.frame(term = ann$GOID, ontology = ann$ONTOLOGY, stringsAsFactors = FALSE)

keep_bp <- term2ont$term[term2ont$ontology == "BP"]

term2gene_bp <- subset(term2gene, term %in% keep_bp)
term2name_bp <- subset(term2name, term %in% keep_bp)

# Sanity check
cat("BP pairs:", nrow(term2gene_bp),
    " unique genes:", length(unique(term2gene_bp$gene)),
    " terms in term2gene_bp:", length(unique(term2gene_bp$term)),
    " terms in term2name_bp:", nrow(term2name_bp), "\n")

# Make sure names and mappings reference the same set of BP terms
terms_common <- intersect(unique(term2gene_bp$term), unique(term2name_bp$term))

term2gene_bp <- subset(term2gene_bp, term %in% terms_common)
term2name_bp <- subset(term2name_bp, term %in% terms_common)

# Gene-set size distribution
summary(as.integer(table(term2gene_bp$term)))
```

## Import differential expression results

```{r, warning=FALSE, comment=NA, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
# One topTable CSV per time point, produced by the DE analysis Rmd -- read in a loop
# rather than as five separately-named objects, since the same processing applies to
# each one regardless of which contrast it is.
toptable_files <- setNames(
  paste0(de_analysis_date, "_Calbicans_toptable_", timepoints, "-Control.csv"),
  timepoints
)

de_results <- lapply(toptable_files, function(f) {
  df <- read.csv(f, stringsAsFactors = FALSE)
  names(df)[names(df) == "X"] <- "GeneID"   # write.csv() default row.names column
  df
})

# Auto-detect column names once per contrast instead of retyping the same five
# detection lines for every comparison.
detect_col <- function(df, candidates) {
  match <- candidates[candidates %in% names(df)]
  if (length(match) == 0) NA_character_ else match[1]
}

col_map <- lapply(de_results, function(df) {
  list(
    id   = detect_col(df, c("GeneID", "ID", "Contig", "Gene")),
    t    = detect_col(df, c("t", "T")),
    padj = detect_col(df, c("adj.P.Val", "adj.P.Val.", "adj_p_val", "adjP")),
    p    = detect_col(df, c("P.Value", "PValue", "pvalue", "P.Val")),
    lfc  = detect_col(df, c("logFC", "log2FoldChange", "LogFC"))
  )
})
```

## Define the universe and differentially expressed gene lists per time point

The universe for each time point is all genes tested in that contrast that also exist in the GO mapping. Significant genes are called at FDR <= 0.05 and split into up/down by the sign of logFC. These numbers are calculated directly from the data below (via `cat()`), rather than typed into the prose, so they can't go stale if the input data changes.

```{r, warning=FALSE, comment=NA, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
fdr_cut <- 0.05

sig_lists <- lapply(timepoints, function(tp) {
  df <- de_results[[tp]]
  cm <- col_map[[tp]]
  df[[cm$id]] <- trimws(df[[cm$id]])

  universe_all <- unique(df[[cm$id]])
  universe     <- intersect(universe_all, unique(term2gene_bp$gene))

  adjOK   <- df[[cm$padj]] <= fdr_cut
  sig_all <- unique(df[[cm$id]][adjOK])

  if (!is.na(cm$lfc)) {
    sig_up   <- unique(df[[cm$id]][adjOK & df[[cm$lfc]] > 0])
    sig_down <- unique(df[[cm$id]][adjOK & df[[cm$lfc]] < 0])
  } else {
    sig_up <- sig_down <- character(0)
  }

  cat(tp, "vs Control -- Universe size:", length(universe),
      " Sig total:", length(sig_all),
      " (Up:", length(sig_up), " Down:", length(sig_down), ")\n")

  list(universe = universe, all = sig_all, up = sig_up, down = sig_down)
})
names(sig_lists) <- timepoints
```

## Over-Representation Analysis (ORA) with clusterProfiler

For ORA, differentially expressed genes (FDR <= 0.05) at each time point are supplied to `clusterProfiler::enricher`, using the custom gene-to-GO mapping (TERM2GENE) built above, with the background restricted to all genes tested in that contrast.

```{r, warning=FALSE, comment=NA, tidy=TRUE, tidy.opts=list(width.cutoff=60)}

run_ora <- function(gene_vec, t2g, t2n, universe) {
  if (length(gene_vec) == 0) return(NULL)
  enricher(
    gene          = intersect(gene_vec, universe),
    universe      = universe,
    TERM2GENE     = t2g,
    TERM2NAME     = t2n,
    pAdjustMethod = "BH",
    qvalueCutoff  = 0.05,
    pvalueCutoff  = 0.05,
    minGSSize     = 5,
    maxGSSize     = 1000
  )
}

ora_results <- lapply(timepoints, function(tp) {
  universe <- sig_lists[[tp]]$universe
  t2g_use  <- subset(term2gene_bp, gene %in% universe)
  t2n_use  <- subset(term2name_bp, term %in% unique(t2g_use$term))

  list(
    all  = run_ora(sig_lists[[tp]]$all,  t2g_use, t2n_use, universe),
    up   = run_ora(sig_lists[[tp]]$up,   t2g_use, t2n_use, universe),
    down = run_ora(sig_lists[[tp]]$down, t2g_use, t2n_use, universe)
  )
})
names(ora_results) <- timepoints
```

```{r, warning=FALSE, comment=NA, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
for (tp in timepoints) {
  for (direction in c("all", "up", "down")) {
    ego <- ora_results[[tp]][[direction]]
    if (is.null(ego)) {
      message(tp, " (", direction, "): no ORA result -- skipping.")
      next
    }

    print(head(as.data.frame(ego)))

    p_ora <- dotplot(ego, showCategory = 20) +
      ggplot2::ggtitle(paste0("GO BP (", tp, " vs Control, ", direction, ")"))
    print(p_ora)
    ggsave(filename = paste0(go_analysis_date, "_Calbicans_GO_ORA_", tp, "_", direction, "_dotplot.pdf"),
           plot = p_ora, width = 8, height = 8, units = "in")

    write.table(as.data.frame(ego),
                file = paste0(go_analysis_date, "_Calbicans_GO_ORA_", tp, "_", direction, ".tsv"),
                sep = "\t", quote = FALSE, row.names = FALSE)
  }
}
```

## Gene Set Enrichment Analysis (GSEA) and cameraPR

Genes are ranked by moderated t-statistic from limma for each time point, and enrichment is computed with `clusterProfiler::GSEA` and `limma::cameraPR`, which accounts for inter-gene correlation.

```{r, warning=FALSE, comment=NA, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
library(limma)

gsea_results   <- list()
camera_results <- list()

for (tp in timepoints) {
  cm       <- col_map[[tp]]
  df       <- de_results[[tp]]
  universe <- sig_lists[[tp]]$universe

  tstats <- df[[cm$t]]
  names(tstats) <- df[[cm$id]]
  keep   <- names(tstats) %in% universe & is.finite(tstats)
  tstats <- sort(tstats[keep], decreasing = TRUE)

  t2g_use <- subset(term2gene_bp, gene %in% universe)
  t2n_use <- subset(term2name_bp, term %in% unique(t2g_use$term))

  gsea_results[[tp]] <- GSEA(
    geneList      = tstats,
    TERM2GENE     = t2g_use,
    TERM2NAME     = t2n_use,
    pAdjustMethod = "BH",
    minGSSize     = 5,
    maxGSSize     = 1000,
    eps           = 0,
    seed          = TRUE
  )

  if (nrow(gsea_results[[tp]]@result) > 0) {
    print(head(as.data.frame(gsea_results[[tp]]@result)))

    p_ridge <- ridgeplot(gsea_results[[tp]], showCategory = 20) +
      ggplot2::ggtitle(paste0(tp, " vs Control"))
    print(p_ridge)
    ggsave(filename = paste0(go_analysis_date, "_Calbicans_GSEA_", tp, "_ridgeplot.pdf"),
           plot = p_ridge, width = 8, height = 8, units = "in")

    p_gsea <- gseaplot2(gsea_results[[tp]], geneSetID = gsea_results[[tp]]@result$ID[1])
    print(p_gsea)
    ggsave(filename = paste0(go_analysis_date, "_Calbicans_GSEA_", tp, "_gseaplot.pdf"),
           plot = p_gsea, width = 8, height = 6, units = "in")

    write.table(as.data.frame(gsea_results[[tp]]@result),
                file = paste0(go_analysis_date, "_Calbicans_GO_GSEA_", tp, ".tsv"),
                sep = "\t", row.names = FALSE, quote = FALSE)
  } else {
    message(tp, ": no GSEA results at current thresholds.")
  }

  gene_sets <- split(t2g_use$gene, t2g_use$term)
  camera_results[[tp]] <- cameraPR(statistic = tstats, index = gene_sets)
  print(head(camera_results[[tp]]))
}
```

## Length-bias sensitivity analysis with goseq

Per-transcript length is extracted once from the shared GTF annotation and reused across every time point, since length depends only on the annotation, not on which contrast is being tested.

```{r, warning=FALSE, comment=NA, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
suppressPackageStartupMessages({
  library(rtracklayer)
  library(GenomicRanges)
  library(goseq)
})

gtf_path <- file.path(reference_dir, "remodeled.gtf")
gtf <- import(gtf_path)

id_field <- if ("transcript_id" %in% colnames(mcols(gtf))) "transcript_id" else {
  f <- grep("transcript", colnames(mcols(gtf)), value = TRUE)
  if (length(f) == 0) stop("No transcript_id-like field in GTF attributes.")
  f[1]
}

if (!any(gtf$type == "exon")) stop("No exon features in GTF.")
exons_all <- gtf[gtf$type == "exon"]

ids       <- as.character(mcols(exons_all)[[id_field]])
split_ex  <- split(exons_all, ids)
tx_len_all <- vapply(split_ex, function(gr) sum(width(reduce(gr))), numeric(1))

goseq_results <- list()

for (tp in timepoints) {
  # The original script reassigned its filtered ID list back to a variable literally
  # named `universe` inside this section -- which, earlier in that script, was the
  # name already used for the interaction-effect universe from a different comparison.
  # That silently overwrote it. Using a loop-local name here (universe_lc) avoids
  # that class of bug entirely.
  universe <- sig_lists[[tp]]$universe
  sig_all  <- sig_lists[[tp]]$all

  lengths <- tx_len_all[universe]
  names(lengths) <- universe

  na_idx <- is.na(lengths)
  if (any(na_idx) && "gene_id" %in% colnames(mcols(gtf))) {
    gene_like <- gtf[gtf$type %in% c("gene", "mRNA", "transcript")]
    gid       <- as.character(mcols(gene_like)[["gene_id"]])
    gene_len  <- vapply(split(gene_like, gid), function(gr) sum(width(reduce(gr))), numeric(1))
    miss      <- names(lengths)[na_idx]
    lengths[miss] <- gene_len[miss]
  }

  keep <- !is.na(lengths)
  if (any(!keep)) {
    warning(tp, ": ", sum(!keep), " universe IDs lack lengths; dropping them for goseq.")
  }
  universe_lc <- universe[keep]
  lengths_lc  <- lengths[keep]

  gene_de <- as.integer(universe_lc %in% sig_all)
  names(gene_de) <- universe_lc

  t2g_use <- term2gene_bp[term2gene_bp$gene %in% universe_lc, , drop = FALSE]
  go_map  <- split(t2g_use$gene, t2g_use$term)

  cat(tp, "vs Control -- Universe:", length(universe_lc),
      " Have lengths for:", sum(!is.na(lengths_lc)), "\n")

  pwf <- nullp(gene_de, bias.data = lengths_lc, plot.fit = FALSE)

  # plotPWF() is base R graphics, not a ggplot object, so ggsave() won't work on it --
  # instead the plot is drawn twice: once into its own PDF device, once normally so it
  # still displays inline (same approach used for the Venn diagram in the DE analysis file).
  pwf_pdf <- paste0(go_analysis_date, "_Calbicans_goseq_PWF_", tp, ".pdf")
  pdf(file = pwf_pdf)
  plotPWF(pwf)
  title(main = paste0(tp, " vs Control"))
  dev.off()

  plotPWF(pwf)
  title(main = paste0(tp, " vs Control"))

  gsq <- goseq(pwf, gene2cat = go_map, use_genes_without_cat = TRUE)
  gsq$padj_over  <- p.adjust(gsq$over_represented_pvalue,  "BH")
  gsq$padj_under <- p.adjust(gsq$under_represented_pvalue, "BH")

  term2name_vec <- setNames(term2name_bp$name, term2name_bp$term)
  gsq$term <- gsq$category
  gsq$name <- unname(term2name_vec[gsq$term])
  gsq$name[is.na(gsq$name)] <- gsq$term

  gsq_out <- gsq[order(gsq$padj_over, gsq$over_represented_pvalue), ]
  write.table(gsq_out,
              file = paste0(go_analysis_date, "_Calbicans_goseq_BP_lengthcorrected_", tp, ".tsv"),
              sep = "\t", row.names = FALSE, quote = FALSE)

  goseq_results[[tp]] <- gsq_out
  print(head(gsq_out[, c("term", "name", "numDEInCat", "numInCat", "over_represented_pvalue", "padj_over")], 20))
}
```

## Session info and package versions

```{r, warning=FALSE, comment=NA, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
sessionInfo()
installed.packages()[, c("Package", "Version")]
write.csv(installed.packages()[, c("Package", "Version")],
          paste0(go_analysis_date, "_installed_packages_versions.csv"), row.names = FALSE)
```
