---
title: "Differential Gene Expression Across a C. albicans Infection Time Course in Planaria"
author: "M. LeGro"
date: "2026-07-21"
output:
  html_document: default
  pdf_document: default
---

## Normalizing and filtering read counts

This picks up from the alignment/`featureCounts` output produced by the alignment Rmd: 18 libraries -- three uninfected controls (CP1-CP3) and five post-infection time points (6hpi, 1dpi, 3dpi, 5dpi, 7dpi) with three replicates each. A `DGEList` is built from the raw counts, normalization factors are calculated, low-count genes are filtered (CPM > 0.5 in at least one sample), and the remaining counts are converted to counts-per-million (CPM).

```{r, warning=FALSE, comment=NA, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
library(tidyverse)   # includes ggplot2, dplyr, readr, etc. -- no need to also load dplyr separately
library(limma)
library(edgeR)
library(RColorBrewer)
library(ComplexHeatmap)
library(circlize)

setwd("/Volumes/PortableSSD/CA_infectionproj/")

analysis_date <- "2026.07.21"                                             # used to label output files
counts_file   <- "2026.07.20_Calbicans_counts_exonfeature.txt"            # written by the alignment Rmd 
# Sample -> condition key, defined once here rather than maintained in a separate
# hand-typed metadata file. The original PTEN script read a standalone meta.txt and
# matched it to the counts matrix by row position; if the two files were ever in a
# different order, samples would be silently mislabeled. Matching by sample_id below
# (rather than trusting row order) removes that risk entirely.
sample_metadata <- data.frame(
  sample_id = c("CP1", "CP2", "CP3",
                "S1", "S2", "S3",
                "S4", "S5", "S6",
                "S7", "S8", "S9",
                "S10", "S11", "S12",
                "S13", "S14", "S15"),
  condition = factor(
    c(rep("Control", 3), rep("6hpi", 3), rep("1dpi", 3),
      rep("3dpi", 3), rep("5dpi", 3), rep("7dpi", 3)),
    levels = c("Control", "6hpi", "1dpi", "3dpi", "5dpi", "7dpi")
  ),
  stringsAsFactors = FALSE
)

# Import raw counts
counts_raw <- read.delim(counts_file, row.names = 1)
countsmat  <- as.matrix(counts_raw[, setdiff(colnames(counts_raw), "Length")])

# Confirm every count column has a matching entry in the sample key, then order the
# metadata to match the counts matrix's actual column order (not the other way around)
stopifnot(all(colnames(countsmat) %in% sample_metadata$sample_id))
meta_ordered <- sample_metadata[match(colnames(countsmat), sample_metadata$sample_id), ]
meta.f <- meta_ordered$condition

# Build the DGEList. Note this is NOT named `list` (as in the original PTEN script) --
# `list` is a base R function, and overwriting it with a DGEList object would silently
# break any later call to list() for the rest of the R session.
dge_list <- DGEList(countsmat, group = meta.f)
dge_list <- calcNormFactors(dge_list)

# Filter out genes with less than 0.5 CPM in every sample
keep     <- rowSums(cpm(dge_list) > 0.5) >= 1
dge_list <- dge_list[keep, ]

CPM <- cpm(dge_list$counts)

write.csv(CPM, file = paste0(analysis_date, "_Calbicans_countsCPM.csv"))
```

After filtering, the dataset went from `r nrow(countsmat)` transcripts to `r nrow(dge_list)` transcripts in the CPM dataset. 

## MDS plot for sample distribution

```{r, warning=FALSE, comment=NA, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
cols <- c("black", "red", "green", "blue", "purple", "orange")
plotMDS(CPM, col = cols[as.numeric(meta.f)])
```

## Fitting the filtered, normalized counts to a linear model

The counts are fit to a linear model with `voomWithQualityWeights` to account for sample-to-sample variation, Bayesian statistics are calculated, and each infected time point is contrasted against the Control group.

```{r, warning=FALSE, comment=NA, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
design <- model.matrix(~0 + meta.f)

# Time point labels start with a digit (6hpi, 1dpi, ...), which isn't a syntactically
# valid R name -- make.names() prefixes those with "X" (6hpi -> X6hpi, etc.), which is
# what makeContrasts requires of the design matrix's column names. This replaces the
# backtick approach, which turned out not to be enough: makeContrasts validates the
# design's column names themselves before it even parses the contrast expressions, so
# quoting the names in the formula doesn't get around it.
colnames(design) <- make.names(levels(meta.f))   # Control, X6hpi, X1dpi, X3dpi, X5dpi, X7dpi

v    <- voomWithQualityWeights(dge_list, design = design, normalize.method = "none", plot = TRUE)
vfit <- lmFit(v, design = design)
vfit <- eBayes(vfit)

# `levels = design` is passed here (the actual design matrix), rather than the raw
# grouping factor as in the original script -- that's the usage limma's docs describe,
# and it removes a fragile dependency on the factor's levels happening to match the
# design's column names exactly.
contrast_matrix <- makeContrasts(
  X6hpi - Control,
  X1dpi - Control,
  X3dpi - Control,
  X5dpi - Control,
  X7dpi - Control,
  levels = design
)

timepoints <- c("6hpi", "1dpi", "3dpi", "5dpi", "7dpi")
colnames(contrast_matrix) <- paste0(timepoints, "_vs_Control")

vfit <- contrasts.fit(vfit, contrasts = contrast_matrix)
vfit <- eBayes(vfit)

results <- decideTests(vfit)

timepoints <- c("6hpi", "1dpi", "3dpi", "5dpi", "7dpi")

# One topTable per contrast, built in a loop instead of five near-identical lines --
# adding or removing a time point later only means editing the `timepoints` vector.
toptables <- setNames(
  lapply(seq_along(timepoints), function(i) topTable(vfit, coef = i, number = Inf, sort.by = "P")),
  timepoints
)
toptable_all <- topTable(vfit, number = Inf)

for (tp in timepoints) {
  write.csv(toptables[[tp]], file = paste0(analysis_date, "_Calbicans_toptable_", tp, "-Control.csv"))
}
write.csv(toptable_all, file = paste0(analysis_date, "_Calbicans_toptable_allcoefs.csv"))
```

*Note: `number = Inf` is used instead of a hard-coded large number (the original used `number = 300000`, a number picked to safely exceed the transcript count) so every gene is returned regardless of dataset size.*

## Calculating z-scores and merging with topTables

```{r, warning=FALSE, comment=NA, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
# scale() computes column-wise statistics, so the matrix is transposed to put genes in
# columns, scaled (z-scored per gene across samples), then transposed back so genes are
# rows and samples are columns again -- same approach as before, just commented so the
# reason for the double-transpose is clear on a re-read.
CPMz  <- scale(t(CPM), center = TRUE, scale = TRUE)
CPMzt <- t(CPMz)

write.csv(CPMzt, file = paste0(analysis_date, "_Calbicans_cpmzscore_all.csv"))

merged <- setNames(
  lapply(timepoints, function(tp) merge.data.frame(toptables[[tp]], CPMzt, by.x = "row.names", by.y = "row.names")),
  timepoints
)

for (tp in timepoints) {
  write.csv(merged[[tp]], file = paste0(analysis_date, "_Calbicans_", tp, "vControl_CPMzscores_wtoptable.csv"))
}
```

## Sorting for significantly differentially expressed transcripts

Genes are considered significant at FDR (adj. P-value) < 0.05. Within that significant set, "up" and "down" are further split by log fold-change beyond +/-1.

```{r, warning=FALSE, comment=NA, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
# In the original script, the "UP"/"DOWN"/"BOTH" subsets were filtered only on logFC,
# without also requiring adj.P.Val < 0.05 -- despite being named with a "_0.05" suffix.
# That means those lists could include genes with a large fold-change but a
# non-significant p-value. Both criteria are applied together here so the "significant"
# label in each list's name actually reflects what's in it. Worth double-checking against
# what you and your coauthors intend before this goes into the manuscript.
sig <- lapply(timepoints, function(tp) {
  df <- merged[[tp]]
  list(
    all  = df[df$adj.P.Val < 0.05, ],
    up   = df[df$adj.P.Val < 0.05 & df$logFC >  1, ],
    down = df[df$adj.P.Val < 0.05 & df$logFC < -1, ],
    both = df[df$adj.P.Val < 0.05 & (df$logFC > 1 | df$logFC < -1), ]
  )
})
names(sig) <- timepoints
```

## Removing extra z-score columns from each timepoint's table

Each merged table above carries z-score columns for all 18 samples, even though a given contrast only needs the Control and matching time point samples. Columns are selected here by sample ID/condition rather than by hard-coded column position.

```{r, warning=FALSE, comment=NA, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
# The original script selected columns with literal indices, e.g. mergedfin[, c(3,4,5,...)].
# That's fragile: if any upstream step changes the number or order of columns, the indices
# silently point at the wrong data with no error. Selecting by name instead is self-documenting
# and safe against upstream changes.
stat_cols <- c("logFC", "AveExpr", "t", "P.Value", "adj.P.Val", "B")

clean_sig_table <- function(df, tp) {
  tp_samples      <- sample_metadata$sample_id[sample_metadata$condition == tp]
  control_samples <- sample_metadata$sample_id[sample_metadata$condition == "Control"]
  out <- df[, c(stat_cols, control_samples, tp_samples)]
  rownames(out) <- df$Row.names
  out
}

for (tp in timepoints) {
  sig[[tp]]$all  <- clean_sig_table(sig[[tp]]$all,  tp)
  sig[[tp]]$up   <- clean_sig_table(sig[[tp]]$up,   tp)
  sig[[tp]]$down <- clean_sig_table(sig[[tp]]$down, tp)
  sig[[tp]]$both <- clean_sig_table(sig[[tp]]$both, tp)
}

```

## Pairwise heatmaps for each time point vs. Control
```{r, warning=FALSE, comment=NA, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
mycolz <- colorRamp2(c(-1.0, 0, 1.0), c("purple", "white", "dark cyan"))

for (tp in timepoints) {
  df <- sig[[tp]]$all   # swap to sig[[tp]]$up / $down / $both for fold-change-thresholded versions

  if (nrow(df) == 0) {
    message(tp, ": no significant genes at FDR < 0.05 -- skipping heatmap.")
    next
  }

  sample_cols <- setdiff(colnames(df), stat_cols)
  mat <- as.matrix(df[, sample_cols])
  storage.mode(mat) <- "numeric"
  rownames(mat) <- make.unique(rownames(mat))

  hm <- Heatmap(mat, name = "Z-score", col = mycolz,
                column_title = paste0(tp, " vs Control"),
                column_names_gp = gpar(fontsize = 8),
                cluster_columns = FALSE, cluster_rows = TRUE)

  print(hm)

  pdf(paste0(analysis_date, "_Calbicans_", tp, "vControl_heatmap.pdf"), width = 6, height = 10)
  draw(hm)
  dev.off()
}
```

## Finding significant genes shared across all time points

```{r, warning=FALSE, comment=NA, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
# The original script found genes shared across all 5 contrasts by chaining pairwise
# merge.data.frame() calls (2 vs 3, then 4 vs 6, then combining those, then adding the
# last one). Besides being hard to follow, repeated merges on tables that share column
# names (logFC, adj.P.Val, etc.) produce auto-suffixed columns like "C01.x.x" that later
# code had to carefully reconstruct. Intersecting gene IDs directly avoids that entirely.
shared_ids <- list(
  all  = Reduce(intersect, lapply(timepoints, function(tp) rownames(sig[[tp]]$all))),
  up   = Reduce(intersect, lapply(timepoints, function(tp) rownames(sig[[tp]]$up))),
  down = Reduce(intersect, lapply(timepoints, function(tp) rownames(sig[[tp]]$down))),
  both = Reduce(intersect, lapply(timepoints, function(tp) rownames(sig[[tp]]$both)))
)

# Per-contrast logFC/adj.P.Val for the shared gene set, with contrast-specific column
# names built directly from the topTables (which are already indexed by gene ID) --
# no merging required, and no suffix collisions to untangle afterward.
build_stats_summary <- function(gene_ids) {
  stats_by_tp <- lapply(timepoints, function(tp) {
    df <- toptables[[tp]][gene_ids, c("logFC", "adj.P.Val")]
    colnames(df) <- paste0(c("logFC_", "adjP_"), tp)
    df
  })
  do.call(cbind, stats_by_tp)
}

stats_summary_all  <- build_stats_summary(shared_ids$all)
stats_summary_both <- build_stats_summary(shared_ids$both)
```

## Cleaning the shared-gene matrices and building supplemental heatmaps

```{r, warning=FALSE, comment=NA, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
zscore_matrix_all  <- as.matrix(CPMzt[shared_ids$all, ])
zscore_matrix_both <- as.matrix(CPMzt[shared_ids$both, ])

write.csv(cbind(stats_summary_all, zscore_matrix_all),
          file = paste0(analysis_date, "_Calbicans_5contrasts_sharedsigDE_FDR0.05.csv"))
write.csv(cbind(stats_summary_both, zscore_matrix_both),
          file = paste0(analysis_date, "_Calbicans_5contrasts_sharedsigDE_LogFC1_FDR0.05.csv"))

mycolz <- colorRamp2(c(-1.0, 0, 1.0), c("purple", "white", "dark cyan"))

# make.unique() guards against duplicate gene IDs (e.g. multi-mapped annotations)
# breaking ComplexHeatmap's row labeling
rownames(zscore_matrix_all)  <- make.unique(rownames(zscore_matrix_all))
rownames(zscore_matrix_both) <- make.unique(rownames(zscore_matrix_both))

hm_all <- Heatmap(zscore_matrix_all, name = "Z-score", col = mycolz,
                   column_names_gp = gpar(fontsize = 8), cluster_columns = FALSE, cluster_rows = TRUE)
hm_both <- Heatmap(zscore_matrix_both, name = "Z-score", col = mycolz,
                    column_names_gp = gpar(fontsize = 8), cluster_columns = FALSE, cluster_rows = TRUE)

hm_all
hm_both

pdf(paste0(analysis_date, "_Calbicans_sharedsigDE_heatmap_all.pdf"), width = 8, height = 10)
draw(hm_all)
dev.off()

pdf(paste0(analysis_date, "_Calbicans_sharedsigDE_heatmap_LogFC1.pdf"), width = 8, height = 10)
draw(hm_both)
dev.off()
```

*Note: the original script wrote these z-score matrices to a `.txt` file and immediately re-read them back in with `read.delim()` before plotting, apparently to force the values to be treated as numeric. That round trip through disk isn't needed -- `as.matrix()` on the already-numeric CPM z-scores above accomplishes the same thing directly in memory.*

## Verifying the statistical significance of dd_smed_v6_2394_0_1 at each timepoint

```{r, warning=FALSE, comment=NA, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
gene_id <- "dd_smed_v6_2394_0_1"

# Was it in the raw counts, and did it survive the low-count filter?
# (these are the same regardless of timepoint, so only need to check once)
grep(gene_id, rownames(countsmat), ignore.case = TRUE, value = TRUE)
grep(gene_id, rownames(dge_list), ignore.case = TRUE, value = TRUE)

# Stats and FDR 0.05 significance for this gene across every timepoint vs Control
gene_report <- lapply(timepoints, function(tp) {
  match_id <- grep(gene_id, rownames(toptables[[tp]]), ignore.case = TRUE, value = TRUE)

  if (length(match_id) == 0) {
    return(data.frame(timepoint = tp, gene_id = NA, logFC = NA,
                       adj.P.Val = NA, significant_FDR0.05 = NA))
  }
  if (length(match_id) > 1) {
    warning(tp, ": multiple gene IDs matched -- check for near-duplicate IDs: ",
            paste(match_id, collapse = ", "))
  }

  stats <- toptables[[tp]][match_id[1], c("logFC", "adj.P.Val")]
  data.frame(
    timepoint = tp,
    gene_id = match_id[1],
    logFC = stats$logFC,
    adj.P.Val = stats$adj.P.Val,
    significant_FDR0.05 = stats$adj.P.Val < 0.05
  )
})

gene_report <- do.call(rbind, gene_report)
gene_report
```
## Creating the venn diagram 

```{r, warning=FALSE, comment=NA, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
dt <- decideTests(vfit, p.value = 0.05, lfc = 1)
summary(dt)

venn_pdf <- paste0(analysis_date, "_Calbicans_VennDiagram_6hpitoDay7.pdf")

venn_plot_args <- list(
  dt,
  include = c("up", "down"),
  counts.col = c("dark cyan", "purple"),
  circle.col = c("red", "blue", "green3", "magenta", "black"),
  mar = rep(1, 4),
  cex = c(1.5, 1.2, 0.5)
)

pdf(file = venn_pdf)
do.call(vennDiagram, venn_plot_args)
dev.off()

# Also draw it inline so it shows up in the knitted document / plot pane
do.call(vennDiagram, venn_plot_args)
```